Interval-based Robust Control of Linear Systems with Generic Costs

### 2020-05-25 main.tex
Remarks from Denis, bis
Conflict merge with overleaf basically wiped my changes... why?

\begin{abstract}
We develop a framework for the adaptive model predictive control (MPC) of a linear system with unknown parameters, in a critical setting where failures are costly and should be prevented at all time.  Through the careful selection and non-trivial adaptation of tools from different communities --namely linear regression, interval prediction and tree-based planning--, we provide an integrated solution to the problem of maximin control with arbitrary bounded costs. We ensured that theoretical guarantees at each layer could be combined to provide what we believe is the first end-to-end analysis in this general setting. The full procedure is designed to produce reasonable and safe behaviours at deployment while progressively recovering an asymptotic optimality. Finally, we relax our modelling assumptions in a multi-model extension based on a data-driven robust model selection mechanism. In addition to our theoretical results we strived to preserve tractable computations for practical applicability, as we illustrate on a simulated application of autonomous driving at a crossroads intersection.\footnote{Source code and videos are available in the Supplementary Material.}
\end{abstract}

@@

Since experiencing failures is out of question, the only way to prevent them from the outset, when the system is first deployed, is to rely on some sort of prior knowledge. In this work, we assume that the system dynamics are partially known, in the form of a parametrized dynamical system with unknown parameters. More precisely, we consider a linear system with state $x\in\Real^p$, acted on by controls $u\in\Real^q$ and perturbations $\omega\in\Real^r$, and following dynamics in the form:
\begin{equation}
\label{eq:dynamics}
\dot{x}(t)=A(\theta)x(t) + B u(t) + D \omega(t),\;t\geq0.
\end{equation}
It is assumed that the parameter vector $\theta$ in the state matrix $A(\theta)\in\Real^{p\times p}$ belongs to a compact set $\Theta \subset \Real^d$. The control matrix $B\in\Real^{p\times q}$ and disturbance matrix $D\in\Real^{p\times r}$ are known. We also assume having access to the observation of $x(t)$ and to a noisy measurement of $\dot{x}(t)$ in the form $y(t)=\dot{x}(t) + C\nu(t)$, where $\nu(t)\in\Real^s$ is a measurement noise and $C\in\Real^{p\times s}$ is known. Assumptions over the disturbance $\omega$ and noise $\nu$ will be detailed further, and we denote $\eta(t) = C\nu(t) + D\omega(t)$. We argue that this structure assumption is realistic given that most industrial applications to date have been relying on physical models to describe their processes and well-engineered controllers to operate them, rather than machine learning. Our framework relaxes this modelling effort by allowing some \emph{structured uncertainty} around the nominal model. We adopt a data-driven scheme to estimate the parameters more accurately as we interact with the true system. Most model-based reinforcement learning algorithms rely on the estimated dynamics to derive the corresponding optimal controls \citep[e.g.][]{Lenz2015,Levine2015}, but suffer from \emph{model bias}: they ignore the error between the learned and true dynamics, which can dramatically degrade control performances \citep{Schneider1997}. %It is particularly likely when maximising an objective under constraints, which naturally pushes the systems to operate in the vicinity of constraint saturation and makes it prone to failure.

@@

The \emph{robust control objective} $V^r$ \citep{Bental2009,Bertsimas2011,Gorissen2015} aims to maximise the worst-case outcome with respect to this confidence region $\cC_{[N],\delta}$:
\begin{equation}
\label{eq:robust-control}
\sup_{\bu\in{(\Real^q)}^\Natural} {V^r(\bu)}, \qquad \text{ where }\qquad {V^r(\bu)} \eqdef \inf_{\substack{A(\theta) \in \cC_{[N],\delta} \\ \bom\in[\underline{\bom},\overline{\bom}]^\Natural}} \left[\sum_{n=N+1}^\infty \gamma^n R(x_n(\bu,\bom))\right],
\end{equation}
$\gamma\in(0,1)$ is a discount factor, and $x_n(\bu,\bom)$ is the state reached at step $n$ under controls $\bu$ and perturbations $\bom$. Maximin problems such as $\eqref{eq:robust-control}$ are notoriously hard in cases such where the reward $R$ has a simple form. With arbitrary functions $R$, we cannot hope to derive an explicit solution.
In our second contribution, we propose a robust MPC algorithm for solving \eqref{eq:robust-control} numerically. In \Cref{sec:prediction}, we leverage recent results from the uncertain system simulation literature to derive an \emph{interval predictor} $[\underline{x}(t),\overline{x}(t)]$ for the system \eqref{eq:dynamics}, illustrated in \Cref{fig:prediction}. This predictor takes the information on the observed current state ${x}_N$, the confidence region $\cC_\delta$, planned control sequence $\bu$ and some admissible bounds on the perturbation $[\underline{\omega}(t),\overline{\omega}(t)]$; and must verify the \emph{inclusion property}:
\begin{equation}
\label{eq:inclusion-property}
\underline{x}(t)\leq x(t)\leq\overline{x}(t), \forall t\geq t_N.
\end{equation}

@@

\paragraph{Linear systems with quadratic costs} Several approaches have been proposed for cumulative regret minimisation in the LQ problem. In the \emph{Optimism in the Face of Uncertainty} paradigm, the best possible dynamics within a high-confidence region is selected under a controllability constraint, to compute the corresponding optimal control in closed-form by solving a Riccati equation. \citep{abbasi-yadkori11a,Ibrahimi2013,Faradonbeh2017} show that this procedures achieves a $\tilde{\cO}\left(N^{1/2}\right)$ regret. Posterior sampling algorithms \citep{Ouyang2017,abeille18a} sample candidate dynamics randomly instead, and obtain the same result. Other works use noise injection for exploration such as \citep{Dean2017,Dean2018}. However, neither optimism nor random exploration fit a critical setting, where ensuring safety requires instead to consider pessimistic outcomes. The work of \citet{Dean2017} is closest to our setting: after an offline estimation phase, they estimate a simple regret between a robust controller and the optimal performance. Our work mainly differs in that it addresses generic costs.

@@

We can enclose this confidence ellipsoid $\eqref{eq:confidence-ellipsoid}$ into a polytope $C_{[N],\delta}$. For simplicity, we present here a simple but coarse strategy: bound the ellipsoid by its enclosing axis-aligned hypercube. We obtain
\begin{align}
    \label{eq:polytope}
     \cC_{[N],\delta} = \left\{ A + \theta_{N,\lambda}^\transp\phi +\sum_{i=1}^{2^d}\alpha_{i}\Delta A_{i}: \alpha\geq 0,  \sum_{i=1}^{2^d}\alpha_{i}=1\right\} \text{ where } h_i\in\{-1,1\}^d,
\end{align}
$\Delta A_{i} = {h_i} \sqrt{\frac{\beta_N(\delta)}{\lambda_{\max}(G_{N,\lambda})}}$. A tighter polytope derivation is presented in the Supplementary Material. % at the price of an increased computational cost required by the diagonalisation of $G_{N,\lambda}$.

@@

\begin{proposition}[Enhanced predictor of \citealt{leurent2019interval}]
\label{prop:predictor}
Assuming that \eqref{eq:polytope} and \Cref{assumpt:metzler} are satisfied for the system \eqref{eq:dynamics}, then the interval predictor following $ \underline{x}(t_N)=\overline{x}(t_N)={x}(t_N)$ and:
\begin{eqnarray}
\dot{\underline{x}}(t) & = & A_{0}\underline{x}(t)-\Delta A_{+}\underline{x}^{-}(t)-\Delta A_{-}\overline{x}^{+}(t)  +Bu(t)+D^{+}\underline{\omega}(t)-D^{-}\overline{\omega}(t),\label{eq:interval-predictor}\\
\dot{\overline{x}}(t) & = & A_{0}\overline{x}(t)+\Delta A_{+}\overline{x}^{+}(t)+\Delta A_{-}\underline{x}^{-}(t)  +Bu(t)+D^{+}\overline{\omega}(t)-D^{-}\underline{\omega}(t),\nonumber
\end{eqnarray}
ensures the inclusion property \eqref{eq:inclusion-property} with confidence level $\delta$..
\end{proposition}

@@

\Cref{fig:predictor_example} compares the stability of the predictors \eqref{eq:predictor-naive} and \eqref{eq:interval-predictor} in a simple example. It suggests to always prefer \eqref{eq:interval-predictor} whenever \Cref{assumpt:metzler} is verified, and only fallback to \eqref{eq:predictor-naive} as a last resort.

@@

\begin{definition}[Surrogate objective]
	Let $\underline{x}_n(\bu), \overline{x}_n(\bu)$ following the dynamics defined in \eqref{eq:interval-predictor} and
\begin{equation}
\label{eq:surrogate-objective} 
\hat{V}^r(\bu) \eqdef \sum_{n=N+1}^\infty \gamma^n \underline{R}_n(\bu)\quad\text{where}\quad\underline{R}_n(\bu) \eqdef \min_{x\in[\underline{x}_n(\bu), \overline{x}_n(\bu)]}  R(x). %\label{eq:pessimistic-rewards}
\end{equation}
\end{definition}
This amounts to changing the reward function, except that the worst case is assessed over the whole past trajectory, which makes this pessimistic reward $\underline{R_n}$ \emph{not Markovian}.

@@

\begin{assumption}[Multi-model ambiguity]
\label{assumpt:multi-model-ambiguity}
The true dynamics $f$ lies within a finite set of candidate models $f^1, \dots, f^M$:
$
\qquad\qquad
\exists m\in[M]: \dot{x}(t) = f^m(x(t), u(t)), \forall t\geq 0.
$
\end{assumption}
We adapt our planning algorithm to balance these concurrent hypotheses in a robust fashion, i.e. maximise a robust objective with discrete ambiguity:
$
%\label{eq:robust-objective-discrete}
V^r = \sup_{a\in\cA^\Natural}\min_{m\in[M]} \sum_{n=N+1}^\infty \gamma^n R_n^m, \refstepcounter{equation}\hfill(\theequation)\label{eq:robust-objective-discrete}
$
where $R_n^m$ is the reward obtained by following the action sequence $a$ up to step $n$ under the dynamics $f^m$.
This objective could be optimised in the same way as in \Cref{sec:control}, but this would result in a coarse and lossy approximation. Instead, we exploit the finite uncertainty structure of \Cref{assumpt:multi-model-ambiguity} to asymptotically recover the true $V^r$ by modifying the \texttt{OPD} algorithm in the following way:

@@

The motivation behind this work is to enable the development of Reinforcement Learning solutions for industrial applications, when it has hitherto been mainly limited to simulated games. In particular, many industries already rely on non-adaptive control systems and could benefit from an increased efficiency, including Oil and Gas, robotics for industrial automation, Data Center cooling, etc. But more often than not, safety-critical constraints bar the use of exploration, and industrials are reluctant to turn to learning-based methods that lack accountability. This work addresses these concerns by focusing on risk-averse decisions and by providing worst-case guarantees. Note however that these guarantees are only as good as the validity of the underlying hypotheses, and \Cref{assumpt:structure} in particular should be submitted to a comprehensive validation procedure; otherwise, decisions formed on a wrong basis could easily lead to dramatic consequences in such critical settings.
Beyond industrial perspectives, this work could be of general interest for risk-averse decision-making. For instance, parametrized epidemiological models have been used to represent the propagation of Covid\nobreakdash-19 and study the impact of several lockdown policies. These model parameters are estimated from observational data and corresponding confidence intervals are often available, but rarely used in the decision-making loop. In contrast, our approach would enable evaluating and optimising the worst-case outcome of such public policies.

>>>>

\begin{abstract}
We develop a framework for the adaptive model predictive control (MPC) of a linear system with unknown parameters, in a critical setting where failures are costly and should be prevented at all time.  Through the careful selection and non-trivial combination of tools from different communities --namely linear regression, interval prediction and tree-based planning--, we provide an integrated solution to the problem of maximin control with arbitrary bounded costs. We ensure that theoretical guarantees at each layer could be combined to provide what we believe is the first end-to-end analysis in this general setting. The full procedure is designed to produce reasonable and safe behaviours, which progressively recover an asymptotic optimality. Finally, we relax our modelling assumptions in a multi-model extension based on a data-driven robust model selection mechanism. In addition to our theoretical results we strived to preserve tractable computations for practical applicability, as we illustrate on a simulated application of autonomous driving at a crossroads intersection.\footnote{Source code and videos are available in the Supplementary Material.}
\end{abstract}

@@

Since experiencing failures is out of question, the only way to prevent them from the outset, when the system is first deployed, is to rely on some sort of prior knowledge. In this work, we assume that the system dynamics are partially known, in the form of a linear differential equation with unknown parameters and inputs. More precisely, we consider a linear system with state $x\in\Real^p$, acted on by controls $u\in\Real^q$ and perturbations $\omega\in\Real^r$, and following dynamics in the form:
\begin{equation}
\label{eq:dynamics}
\dot{x}(t)=A(\theta)x(t) + B u(t) + D \omega(t),\;t\geq0,
\end{equation}
where the parameter vector $\theta$ in the state matrix $A(\theta)\in\Real^{p\times p}$ belongs to a compact set $\Theta \subset \Real^d$. The control matrix $B\in\Real^{p\times q}$ and disturbance matrix $D\in\Real^{p\times r}$ are known. We also assume having access to the observation of $x(t)$ and to a noisy measurement of $\dot{x}(t)$ in the form $y(t)=\dot{x}(t) + C\nu(t)$, where $\nu(t)\in\Real^s$ is a measurement noise and $C\in\Real^{p\times s}$ is known. Assumptions over the disturbance $\omega$ and noise $\nu$ will be detailed further, and we denote $\eta(t) = C\nu(t) + D\omega(t)$. We argue that this structure assumption is realistic given that most industrial applications to date have been relying on physical models to describe their processes and well-engineered controllers to operate them, rather than machine learning. Our framework relaxes this modelling effort by allowing some \emph{structured uncertainty} around the nominal model. We adopt a data-driven scheme to estimate the parameters more accurately as we interact with the true system. Most model-based reinforcement learning algorithms rely on the estimated dynamics to derive the corresponding optimal controls \citep[e.g.][]{Lenz2015,Levine2015}, but suffer from \emph{model bias}: they ignore the error between the learned and true dynamics, which can dramatically degrade control performances \citep{Schneider1997}. %It is particularly likely when maximising an objective under constraints, which naturally pushes the systems to operate in the vicinity of constraint saturation and makes it prone to failure.

@@

The \emph{robust control objective} $V^r$ \citep{Bental2009,Bertsimas2011,Gorissen2015} aims to maximise the worst-case outcome with respect to this confidence region $\cC_{[N],\delta}$:
\begin{equation}
\label{eq:robust-control}
\sup_{\bu\in{(\Real^q)}^\Natural} {V^r(\bu)}, \qquad \text{ where }\qquad {V^r(\bu)} \eqdef \inf_{\substack{A(\theta) \in \cC_{[N],\delta} \\ \bom\in[\underline{\bom},\overline{\bom}]^\Natural}} \left[\sum_{n=N+1}^\infty \gamma^n R(x_n(\bu,\bom))\right],
\end{equation}
$\gamma\in(0,1)$ is a discount factor, and $x_n(\bu,\bom)$ is the state reached at step $n$ under controls $\bu$ and perturbations $\bom$ within the given admissible bounds $[\underline\omega(t),\overline\omega(t)]$. Maximin problems such as $\eqref{eq:robust-control}$ are notoriously hard if the reward $R$ has a simple form. With arbitrary functions $R$, we cannot hope to derive an explicit solution.
In our second contribution, we propose a robust MPC algorithm for solving \eqref{eq:robust-control} numerically. In \Cref{sec:prediction}, we leverage recent results from the uncertain system simulation literature to derive an \emph{interval predictor} $[\underline{x}(t),\overline{x}(t)]$ for the system \eqref{eq:dynamics}, illustrated in \Cref{fig:prediction}. This predictor takes the information on the observed current state ${x}_N$, the confidence region $\cC_\delta$, planned control sequence $\bu$ and admissible perturbation bounds $[\underline{\omega}(t),\overline{\omega}(t)]$; and must verify the \emph{inclusion property}:
\begin{equation}
\label{eq:inclusion-property}
\underline{x}(t)\leq x(t)\leq\overline{x}(t), \forall t\geq t_N.
\end{equation}

@@

\paragraph{Linear systems with quadratic costs} Several approaches have been proposed for cumulative regret minimisation in the LQ problem. In the \emph{Optimism in the Face of Uncertainty} paradigm, the best possible dynamics within a high-confidence region is selected under a controllability constraint, to compute the corresponding optimal control in closed-form by solving a Riccati equation. \citep{abbasi-yadkori11a,Ibrahimi2013,Faradonbeh2017} show that this procedure achieves a $\tilde{\cO}\left(N^{1/2}\right)$ regret. Posterior sampling algorithms \citep{Ouyang2017,abeille18a} select candidate dynamics randomly instead, and obtain the same result. Other works use noise injection for exploration such as \citep{Dean2017,Dean2018}. However, neither optimism nor random exploration fit a critical setting, where ensuring safety requires instead to consider pessimistic outcomes. The work of \citet{Dean2017} is closest to our setting: after an offline estimation phase, they estimate a simple regret between a robust controller and the optimal performance. Our work mainly differs in that it addresses generic costs.

@@

We can enclose this confidence ellipsoid $\eqref{eq:confidence-ellipsoid}$ into a polytope $C_{[N],\delta}$. For simplicity, we present here a simple but coarse strategy: bound the ellipsoid by its enclosing axis-aligned hypercube. We obtain
\begin{align}
    \label{eq:polytope}
     \cC_{[N],\delta} = \left\{ A_0 +\sum_{i=1}^{2^d}\alpha_{i}\Delta A_{i}: \alpha\geq 0,  \sum_{i=1}^{2^d}\alpha_{i}=1\right\} \text{ where } A_0 = A + \theta_{N,\lambda}^\transp\phi,\; h_i\in\{-1,1\}^d,
\end{align}
$\Delta A_{i} = {h_i} \sqrt{\frac{\beta_N(\delta)}{\lambda_{\max}(G_{N,\lambda})}}$. A tighter polytope derivation is presented in the Supplementary Material. % at the price of an increased computational cost required by the diagonalisation of $G_{N,\lambda}$.

@@

\begin{proposition}[Enhanced predictor of \citealt{leurent2019interval}]
\label{prop:predictor}
Assuming that \eqref{eq:polytope} and \Cref{assumpt:metzler} are satisfied for the system \eqref{eq:dynamics}, then the interval predictor following $ \underline{x}(t_N)=\overline{x}(t_N)={x}(t_N)$ and:
\begin{eqnarray}
\dot{\underline{x}}(t) & = & A_{0}\underline{x}(t)-\Delta A_{+}\underline{x}^{-}(t)-\Delta A_{-}\overline{x}^{+}(t)  +Bu(t)+D^{+}\underline{\omega}(t)-D^{-}\overline{\omega}(t),\label{eq:interval-predictor}\\
\dot{\overline{x}}(t) & = & A_{0}\overline{x}(t)+\Delta A_{+}\overline{x}^{+}(t)+\Delta A_{-}\underline{x}^{-}(t)  +Bu(t)+D^{+}\overline{\omega}(t)-D^{-}\underline{\omega}(t),\nonumber
\end{eqnarray}
ensures the inclusion property \eqref{eq:inclusion-property} with confidence level $\delta$.
\end{proposition}

@@

\Cref{fig:predictor_example} compares the performance of the predictors \eqref{eq:predictor-naive} and \eqref{eq:interval-predictor} in a simple example. It suggests to always prefer \eqref{eq:interval-predictor} whenever \Cref{assumpt:metzler} is verified, and only fallback to \eqref{eq:predictor-naive} as a last resort.

@@

\begin{definition}[Surrogate objective]
	Let $\underline{x}_n(\bu), \overline{x}_n(\bu)$ following the dynamics defined in \eqref{eq:interval-predictor} and
\begin{equation}
\label{eq:surrogate-objective} 
\hat{V}^r(\bu) \eqdef \sum_{n=N+1}^\infty \gamma^n \underline{R}_n(\bu)\quad\text{where}\quad\underline{R}_n(\bu) \eqdef \min_{x\in[\underline{x}_n(\bu), \overline{x}_n(\bu)]}  R(x). %\label{eq:pessimistic-rewards}
\end{equation}
\end{definition}
Such a substitution makes this pessimistic reward $\underline{R_n}$ \emph{not Markovian}, since the worst case is assessed over the whole past trajectory.

@@

\begin{assumption}[Multi-model ambiguity]
\label{assumpt:multi-model-ambiguity}
The true dynamics $f$ lies within a finite set of candidate models $f^1, \dots, f^M$:
$
\qquad\qquad
\exists m\in[M]: \dot{x}(t) = f^m(x(t), u(t)), \forall t\geq 0.
$
\end{assumption}
We adapt our planning algorithm to balance these concurrent hypotheses in a robust fashion, i.e. maximise a robust objective with discrete ambiguity:
\begin{equation}
\label{eq:robust-objective-discrete}
V^r = \sup_{a\in\cA^\Natural}\min_{m\in[M]} \sum_{n=N+1}^\infty \gamma^n R_n^m, 
%\refstepcounter{equation}\hfill(\theequation)\label{eq:robust-objective-discrete}
\end{equation}
where $R_n^m$ is the reward obtained by following the action sequence $a$ up to step $n$ under the dynamics $f^m$.
This objective could be optimised in the same way as in \Cref{sec:control}, but this would result in a coarse and lossy approximation. Instead, we exploit the finite uncertainty structure of \Cref{assumpt:multi-model-ambiguity} to asymptotically recover the true $V^r$ by modifying the \texttt{OPD} algorithm in the following way:

@@

The motivation behind this work is to enable the development of Reinforcement Learning solutions for industrial applications, when it has been mainly limited to simulated games so far. In particular, many industries already rely on non-adaptive control systems and could benefit from an increased efficiency, including Oil and Gas, robotics for industrial automation, Data Center cooling, etc. But more often than not, safety-critical constraints bar the use of exploration, and industrials are reluctant to turn to learning-based methods that lack accountability. This work addresses these concerns by focusing on risk-averse decisions and by providing worst-case guarantees. Note however that these guarantees are only as good as the validity of the underlying hypotheses, and \Cref{assumpt:structure} in particular should be submitted to a comprehensive validation procedure; otherwise, decisions formed on a wrong basis could easily lead to dramatic consequences in such critical settings.
Beyond industrial perspectives, this work could be of general interest for risk-averse decision-making. For instance, parametrized epidemiological models have been used to represent the propagation of Covid\nobreakdash-19 and study the impact of several lockdown policies. These model parameters are estimated from observational data and corresponding confidence intervals are often available, but rarely used in the decision-making loop. In contrast, our approach would enable evaluating and optimising the worst-case outcome of such public policies.
